Management induced changes of soil organic carbon on global croplands

### 2020-12-14 SOCbudgetPaper/methods.Rmd

In Sect. \@ref(sec:carbonbudget) we introduce the basic concept of SOC dynamics as applied in this study and explained in more detail within the refinement of the IPCC guidelines vol. 4 ([@calvo_buendia_ipcc_2019]). We additionally describe how we configured and extended the steady-state method (for model code see [@karstens_mrsoil_2020]). In Sect. \@ref(sec:tier1) we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines ([@eggleston_hs_ipcc_2006], [@calvo_buendia_ipcc_2019]). 
Section \@ref(sec:agrimanagement) provides a detailed description of the global, gridded management data used to drive the model, including crop production levels, residue input rates, manure amendments, and the adoption of irrigation and tillage practices (for model code see [@bodirsky_mrcommons_2020]). In Sect. \@ref(sec:scenarios) we define the scenarios used to complement our historic model results.

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks $SOC^{\mathrm{eq}}$ are given by 
\begin{equation}
SOC^{\mathrm{eq}}_{i,t,sub,lu} =\frac{C^{\mathrm{in}}_{i,t,sub,lu}}{k_{i,t,sub,lu}}
(\#eq:inoutflow)
\end{equation}
with $C^{\textrm{in}}$ being the carbon inputs to the soil, $k$ denotes the soil organic carbon decay rate. This equation is valid for all grid cells $i$ and all years $t$. We use the steady-state method for our calculations, which assumes three soil carbon sub-pools $sub$ (active, slow and passive) and interactions between them, following the approach in the Century model [@parton_analysis_1987]). Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are land-use type $lu$ specific. 
We distinguish two land-use types: cropland and uncropped land under potential natural vegetation as representative for all other land-use types including forestry and pastures (referred to as natural vegetation in the following).

@@

  ```{=latex}
    \begin{table*}[h]
    \caption{Type and data sources for carbon inputs and parameterization to different land-use types }
    \begin{tabular}{p{0.15\textwidth} p{0.20\textwidth} p{0.25\textwidth} p{0.25\textwidth}}
    \tophline
    \textbf{land-use types}   & \textbf{source of carbon inputs} & \textbf{data source} & \textbf{nitrogen and lignin content} \\
    \middlehline
    \multirow{3}{*}{Cropland} & above-ground residues & \multirow{3}{*}{\begin{minipage}[t]{0.28\columnwidth}\raggedright
                                                                        \cite{faostat_faostat_2016}, \\
                                                                        \cite{schaphoff_lpjml4_2018} (LPJmL),\\
                                                                        \cite{weindl_livestock_2017} \end{minipage}} & 
                                                        \multirow{2}{*}{\begin{minipage}[t]{0.28\columnwidth}\raggedright
                                                                        LG:C generic values according to Table 5.5B\\
                                                                        in \cite{calvo_buendia_ipcc_2019},\\
                                                                        crop-type specific N:C-ratios from
                                                                        \cite{bodirsky_mrcommons_2020}\end{minipage}} \\ \cline{2-2}
                              & below-ground residues &  &  \\ \cline{2-2}
                              & manure                &  & Table 5.5C in \cite{calvo_buendia_ipcc_2019} \\ 
                              \hline
    Natural vegetation        & annual litterfall &  \cite{schaphoff_lpjml4_2018} (LPJmL) 
                              & \begin{minipage}[t]{0.28\columnwidth}\raggedright\strut 
                                  \cite{calvo_buendia_ipcc_2019} (CENTURY parameterization) \strut \end{minipage}\tabularnewline
   \bottomhline
   \end{tabular}
   \label{tab:datasourceinputs}
   \belowtable{}
   \end{table*}
```

@@

Note that the absolute changes $\Delta SOC$ can be also interpreted as the SOC debt ([@sanderman_soil_2017]) due to human cropping activities; whereas relative changes $F^{\mathrm{SCF}}$ can be considered stock change factors as defined within the IPCC guidelines of 2006 [@eggleston_hs_ipcc_2006]. The SOC debt is equivalent to the negated cumulative SOC component of  of human land use emission ([@pugh_simulated_2015]).

@@

Following the IPCC guidelines, we split the harvest index into a production ($HI^{\mathrm{prod}}$) and an area dependent ($HI^{\mathrm{area}}$) fraction [@eggleston_hs_ipcc_2006]. Deviating from [@bodirsky_n2o_2012] we use harvested instead of physical crop area to account for increased residue biomass due to multiple cropping and decreased residue amounts due to fallow land. We assume that all $BGR$ are left in the soil, whereas $AGR$ can be burned or harvested for other purposes such as feeding animals [@weindl_livestock_2017], fuel or for material use.

A country-specific fixed share of the $AGR$ is assumed to be burned on field depending on the per-capita income of the country. Following [@smil_nitrogen_1999]) we assume a burn share of $25\%$ for low-income countries according to World Bank definitions ($<\,1000\,\tfrac{USD}{yr\,cap}$), 15% for high-income  ($>\,10000\,\tfrac{USD}{yr\,cap}$) and linearly interpolate shares for all middle-income countries depending on their per-capita income. Depending on the crop group, 80–90% of the carbon in the crop residues burned in the fields is lost within the combustion process [@eggleston_hs_ipcc_2006].

@@

To transform dry matter estimates into carbon and nitrogen, we compiled crop-group and plant-part specific carbon resp. nitrogen to dry matter (c/dm, n/dm) ratios (see \@ref(append:Tablec2dm)).

@@

Due to missing information about the litterfall composition, we assume woody potential natural vegetation for the lignin-to-carbon (short LG:C) and nitrogen-to-carbon (short N:C) ratios for all land under natural vegetation. This is however a strong simplification. Therefore, we test the sensitivity of the model for other natural vegetation parameterizations. Perennial grasses (scenario name \textit{LitterPNV-PerennialGrasses}) are given by the IPCC guidelines ([@calvo_buendia_ipcc_2019]), but for woody biomass we used the CENTURY configuration file ([@century_model_2000]) for trees as parameterization were not given. Whereas LG:C ratios ranges between $0.09-0.35$ with an average of $0.24$ (default value) for different tree compartments and different tree types, N:C ratios range between $0.001-0.05$ with an average of $0.015$ (default value). We decided to average over all tree compartments and tree types equally for the default values, since we have no data available on tree type distribution and tree compartment composition. We test however a range of different plant types as specified in Table @ref(tab:scenlitterpnv).

>>>>

In Sect. \@ref(sec:carbonbudget) we introduce the basic concept of SOC dynamics as applied in this study and explained in more detail within the refinement of the IPCC guidelines vol. 4 ([@calvo_buendia_ipcc_2019]). We additionally describe how we configured and extended the steady-state method (for model code see [@karstens_mrsoil_2020]). In Sect. \@ref(sec:tier1) we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines ([@eggleston_ipcc_2006], [@calvo_buendia_ipcc_2019]). 
Section \@ref(sec:agrimanagement) provides a detailed description of the global, gridded management data used to drive the model, including crop production levels, residue input rates, manure amendments, and the adoption of irrigation and tillage practices (for model code see [@bodirsky_mrcommons_2020]). In Sect. \@ref(sec:scenarios) we define the scenarios used to complement our historic model results.

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks $SOC^{\mathrm{eq}}$ are given by 
\begin{equation}
SOC^{\mathrm{eq}}_{i,t,sub,lu} =\frac{C^{\mathrm{in}}_{i,t,sub,lu}}{k_{i,t,sub,lu}}
(\#eq:inoutflow)
\end{equation}
with $C^{\textrm{in}}$ being the carbon inputs to the soil, $k$ denotes the soil organic carbon decay rate. This equation is valid for all grid cells $i$ and all years $t$. We use the steady-state method for our calculations, which assumes three soil carbon sub-pools $sub$ (active, slow and passive) and interactions between them, following the approach in the Century model ([@parton_analysis_1987]). Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are land-use type $lu$ specific. 
We distinguish two land-use types: cropland and uncropped land under potential natural vegetation as representative for all other land-use types including forestry and pastures (referred to as natural vegetation in the following).

@@

  ```{=latex}
    \begin{table*}[h]
    \caption{Type and data sources for carbon inputs and parameterization to different land-use types }
    \begin{tabular}{p{0.15\textwidth} p{0.20\textwidth} p{0.25\textwidth} p{0.40\textwidth}}
    \tophline
    \textbf{land-use types}   & \textbf{source of carbon inputs} & \textbf{data source} & \textbf{nitrogen and lignin content} \\
    \middlehline
    \multirow{3}{*}{Cropland} & above-ground residues & \multirow{3}{*}{\begin{minipage}[t]{1\columnwidth}\raggedright
                                                                        \cite{faostat_faostat_2016}, \\
                                                                        \cite{schaphoff_lpjml4_2018} (LPJmL),\\
                                                                        \cite{weindl_livestock_2017} \end{minipage}} & 
                                                        \multirow{3}{*}{\begin{minipage}[t]{1\columnwidth}\raggedright
                                                                        LG:C generic values according to Table 5.5B\\
                                                                        in \cite{calvo_buendia_ipcc_2019},\\
                                                                        crop-type specific N:C-ratios from
                                                                        \cite{bodirsky_mrcommons_2020}\end{minipage}} \\
                              & below-ground residues &  &  \\
                              & manure                &  &  \\ 
                              \hline
    Natural vegetation        & annual litterfall &  \cite{schaphoff_lpjml4_2018} (LPJmL) 
                              & \begin{minipage}[t]{0.28\columnwidth}\raggedright\strut 
                                  \cite{calvo_buendia_ipcc_2019} (CENTURY parameterization) \strut \end{minipage}\tabularnewline
   \bottomhline
   \end{tabular}
   \label{tab:datasourceinputs}
   \belowtable{}
   \end{table*}
```

@@

Note that the absolute changes $\Delta SOC$ can be also interpreted as the SOC debt ([@sanderman_soil_2017]) due to human cropping activities; whereas relative changes $F^{\mathrm{SCF}}$ can be considered stock change factors as defined within the IPCC guidelines of 2006 [@eggleston_ipcc_2006]. Moreover, $\Delta SOC$ is equivalent to the negated cumulative SOC component of human land-use change emissions ([@pugh_simulated_2015]).

@@

Following the IPCC guidelines, we split the harvest index into a production ($HI^{\mathrm{prod}}$) and an area dependent ($HI^{\mathrm{area}}$) fraction [@eggleston_ipcc_2006]. Deviating from [@bodirsky_n2o_2012] we use harvested instead of physical crop area to account for increased residue biomass due to multiple cropping and decreased residue amounts due to fallow land. We assume that all $BGR$ are left in the soil, whereas $AGR$ can be burned or harvested for other purposes such as feeding animals [@weindl_livestock_2017], fuel or for material use.

A country-specific fixed share of the $AGR$ is assumed to be burned on field depending on the per-capita income of the country. Following [@smil_nitrogen_1999]) we assume a burn share of $25\%$ for low-income countries according to World Bank definitions ($<\,1000\,\tfrac{USD}{yr\,cap}$), 15% for high-income  ($>\,10000\,\tfrac{USD}{yr\,cap}$) and linearly interpolate shares for all middle-income countries depending on their per-capita income. Depending on the crop group, 80–90% of the carbon in the crop residues burned in the fields is lost within the combustion process [@eggleston_ipcc_2006].

@@

To transform dry matter estimates into carbon and nitrogen, we compiled crop-group and plant-part specific carbon resp. nitrogen to dry matter (c/dm, n/dm) ratios (see \@ref(tab:c2dm)).

@@

Due to missing information about the litterfall composition, we assume woody potential natural vegetation for the lignin-to-carbon (short LG:C) and nitrogen-to-carbon (short N:C) ratios for all land under natural vegetation. This is however a strong simplification. Therefore, we test the sensitivity of the model for other natural vegetation parameterizations. Perennial grasses (scenario name \textit{LitterPNV-PerennialGrasses}) are given by the IPCC guidelines ([@calvo_buendia_ipcc_2019]), but for woody biomass we used the CENTURY configuration file ([@century_model_2000]) for trees as parameterization were not given. Whereas LG:C ratios ranges between $0.09-0.35$ with an average of $0.24$ (default value) for different tree compartments and different tree types, N:C ratios range between $0.001-0.05$ with an average of $0.015$ (default value). We decided to average over all tree compartments and tree types equally for the default values, since we have no data available on tree type distribution and tree compartment composition. We test however a range of different plant types as specified in Table \@ref(tab:scenlitterpnv).

### 2020-12-14 SOCbudgetPaper/results.Rmd

```{r captionSOCmaps, echo=FALSE}
captionSOCmaps <- "(a): Distribution of total global SOC stocks for the first 30 cm on cropland shows high carbon stocks in high yielding areas. (b)+(c): Absolute (b) and relative (c) SOC stocks changes compared to a potential natural state showing different hot spots of SOC dynamics. Whereas absolute losses $\\Delta SOC$ are often highest in temperate dry regions, relative losses $F^\\mathrm{SCF}$ are often larger in tropical moist areas. (d): The SOC debt between SOC under historic land use and potential natural vegetation is decreasing over time, meaning net SOC gains on global croplands over the last decades. "
```

@@

Focusing of the temporal dimension of SOC stock changes (see Fig. \@ref(fig:SOCmaps)(c)), global SOC debt has decreased around 15% over the last couple of decades to $22\unit{GtC}$. This leads to a sequestration rate of $0.11\unit{GtC yr^{-1}}$ and considering our estimate of the global SOC stock of $675\unit{GtC}$ in 2010 to $0.2$ per mille per year.

## Agricultural management effects on SOC debt

Figure \@ref(fig:SOCscen)(a) reveals the relative impact of management effects by keeping carbon inflows from crop residue returning or manure application constant at the level of 1975 and neglecting the adoption of no-tillage practices. Without changes in management regimes, especially in C inflows from crop residues to the soil, global SOC debt on cropland would be increase at a rate of $0.1\unit{GtC yr^{-1}}$. The constResidue scenario dominates the effect and is only single-off scenario that still losses SOC compared to natural vegetation at a rate of $0.06\unit{GtC yr^{-1}}$. Whereas both the constManure and constTillage scenarios, show only small deviations from the historical values (sequestration rate of $0.09\unit{GtC yr^{-1}}$ for both), the effect of no-tillage gets stronger till 2010. 
The strong impact of almost doubling C inputs from crop residue biomass over a period of 35 years is visible in Fig. \@ref(fig:SOCscen)(b) as well.

@@

```{r SOCscen, out.width = "18cm", fig.cap = captionSOCscen, echo=FALSE, fig.pos="h"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/scenario_horiz.png")
```

The results of the sensitivity analysis concerning the initialization of SOC stocks at the beginning of the spin-up phase are shown in Fig. \@ref(fig:SOcinit). The Initial-natveg scenario initializes the start year with the steady-state SOC under potential natural vegetation for all land-use types, compared to the default assumption (Initial-lu), which assumes landuse-type specific steady-state SOC stocks in 1901. The $\Delta SOC$ estimation almost halves (from ~26 GtC to ~14 GtC) as well as the decrease in of $\Delta SOC$ for over the last 35 years (from ~4 GtC to ~2.5 GtC) for the Inital-natveg scenario.

@@

```{r SOCinit, out.width = "18cm", fig.cap = captionSOCinit, echo=FALSE, fig.pos="H"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/scenario_init.png")
```
 Figure \@ref(fig:)

@@

```{r FlowFig, out.width = "16cm", fig.cap = caption_FlowFig, echo=FALSE, fig.pos="h"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/CarbonBudget.png")
```

<!-- ```{r table2, echo = FALSE}
# table <- readRDS("../ResultNotebooks/Output/SFC_comparison.rds")
# knitr::kable(table$comparison_table, "latex", digit=2, 
#              caption="$F^{SOC}$ in comparison to IPCC Tier 1 default factors: Stock change factors are in good agreement with the default values of the IPCC in general. For the tropical regions the assumptions changed notablly from the guidelines in 2006 to the update in 2019. leaving our results in very good agreement with the old default assumptions. Default assumption are given under the assumption of medium input systems, which, considering the yield gap in mainly developing regions in the tropics, might be an overestimation and decrease $F^{SOC}$ by additional 5-8 percent. Also: modelled $F^{SOC}$ have increased for all climates over time .")
# ```
# ```{r table, echo=FALSE}
# table <- readRDS("../ResultNotebooks/Output/SFC_climatezone.rds")
# knitr::kable(table$naturalstate_table[3:10,], "latex", digit=2, 
#              caption="S)
# ``` -->

>>>>

```{r captionSOCmaps, echo=FALSE}
captionSOCmaps <- "(a): Distribution of total global SOC stocks for the first 30 cm on cropland shows high carbon stocks in high yielding areas. (b)+(c): Absolute (b) and relative (c) SOC stocks changes compared to a potential natural state showing different hot spots of SOC dynamics. Whereas absolute losses $\\Delta SOC$ are often highest in temperate dry regions, relative losses $F^\\mathrm{SCF}$ are often larger in tropical moist areas. (d): $\\Delta SOC$ between SOC under historic land use and potential natural vegetation is decreasing over time, meaning net SOC gains on global croplands over the last decades. "
```

@@

Focusing of the temporal dimension of SOC stock changes (see Fig. \@ref(fig:SOCmaps)(c)), global $\Delta SOC$  has decreased around 15% over the last couple of decades to $22\unit{GtC}$. This leads to a sequestration rate of $0.11\unit{GtC yr^{-1}}$ and considering our estimate of the global SOC stock of $675\unit{GtC}$ in 2010 to $0.2$ per mille per year.

## Agricultural management effects on $\Delta SOC$

Figure \@ref(fig:SOCscen)(a) reveals the relative impact of management effects by keeping carbon inflows from crop residue returning or manure application constant at the level of 1975 and neglecting the adoption of no-tillage practices. Without changes in management regimes, especially in C inflows from crop residues to the soil, global $\Delta SOC$  on cropland would be increase at a rate of $0.1\unit{GtC yr^{-1}}$. The constResidue scenario dominates the effect and is only single-off scenario that still losses SOC compared to natural vegetation at a rate of $0.06\unit{GtC yr^{-1}}$. Whereas both the constManure and constTillage scenarios, show only small deviations from the historical values (sequestration rate of $0.09\unit{GtC yr^{-1}}$ for both), the effect of no-tillage gets stronger till 2010. 
The strong impact of almost doubling C inputs from crop residue biomass over a period of 35 years is visible in Fig. \@ref(fig:SOCscen)(b) as well.

@@

```{r SOCscen, out.width="14cm", fig.cap = captionSOCscen, echo=FALSE}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/scenario_horiz.png")
```

The results of the sensitivity analysis concerning the initialization of SOC stocks at the beginning of the spin-up phase are shown in Fig. \@ref(fig:SOCinit). The Initial-natveg scenario initializes the start year with the steady-state SOC under potential natural vegetation for all land-use types, compared to the default assumption (Initial-lu), which assumes landuse-type specific steady-state SOC stocks in 1901. The $\Delta SOC$ estimation almost halves (from $~26\unit{GtC}$ to $~14\unit{GtC}$) as well as the decrease in of $\Delta SOC$ for over the last 35 years (from $~4\unit{GtC}$ to $~2.5\unit{GtC}$) for the Inital-natveg scenario.

@@

```{r SOCinit, out.width="10cm", fig.cap = captionSOCinit, echo=FALSE}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/scenario_init.png")
```

@@

```{r FlowFig, out.width = "16cm", fig.cap = caption_FlowFig, echo=FALSE}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/CarbonBudget.png")
```
